Faithful Model Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-12-13 experiments/notebooks/poc.qmd
acknowledgements

```{julia}
include("$(pwd())/notebooks/setup.jl")
eval(setup_notebooks)
```

```{julia}
# Counteractual data:
n_obs = 1000
counterfactual_data = load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.))
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
batch_size = Int(round(nobs/10))
epochs = 100
```

@@

```{julia}
𝒟x = Normal()
𝒟y = Categorical(ones(2) ./ 2)
sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=50)
n_hidden = 16
clf = JointEnergyClassifier(
    sampler;
    builder=MLJFlux.MLP(
        hidden=(n_hidden, n_hidden, n_hidden), 
        σ=Flux.swish
    ),
    batch_size=batch_size,
    finalizer=Flux.softmax,
    loss=Flux.Losses.crossentropy,
    jem_training_params=(
        α=[1.0,1.0,1e-1],
        verbosity=10,
    ),
    epochs=epochs,
    sampling_steps=30,
)
```

@@

```{julia}
#| echo: false
niter = 1000
jem = mach.model.model.jem
batch_size = mach.model.model.batch_size
X = Float32.(matrix(X))
if typeof(jem.sampler) <: ConditionalSampler

    plts = []
    for target in 1:2
        X̂ = generate_conditional_samples(jem, batch_size, target; niter=niter) 
        ex = extrema(hcat(X,X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        plt = Plots.contour(
            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
    display(plt)
end
```

>>>>

```{julia}
using Pkg; Pkg.activate("experiments")
include("$(pwd())/experiments/setup_env.jl")
using TaijaPlotting
```

```{julia}
# Counteractual data:
n_obs = 1000
counterfactual_data = load_blobs(n_obs; cluster_std=0.1, center_box=(-1. => 1.))
X = counterfactual_data.X
y = counterfactual_data.y
labels = counterfactual_data.output_encoder.labels
input_dim, nobs = size(X)
_batch_size = Int(round(nobs/10))
epochs = 100
```

@@

```{julia}
𝒟x = Normal()
𝒟y = Categorical(ones(2) ./ 2)
sampler = ConditionalSampler(𝒟x, 𝒟y, input_size=size(X)[1:end-1], batch_size=50)
n_hidden = 16
clf = JointEnergyClassifier(
    sampler;
    builder=MLJFlux.MLP(
        hidden=(n_hidden, n_hidden, n_hidden), 
        σ=Flux.swish
    ),
    batch_size=_batch_size,
    finaliser=Flux.softmax,
    loss=Flux.Losses.crossentropy,
    jem_training_params=(
        α=[1.0,1.0,1e-1],
        verbosity=10,
    ),
    epochs=epochs,
    sampling_steps=30,
)
```

@@

```{julia}
#| echo: false
niter = 1000
jem = mach.model.model.jem
_batch_size = mach.model.model.batch_size
X = Float32.(matrix(X))
if typeof(jem.sampler) <: ConditionalSampler

    plts = []
    for target in 1:2
        X̂ = generate_conditional_samples(jem, _batch_size, target; niter=niter) 
        ex = extrema(hcat(X,X̂), dims=2)
        xlims = ex[1]
        ylims = ex[2]
        x1 = range(1.0f0.*xlims...,length=100)
        x2 = range(1.0f0.*ylims...,length=100)
        plt = Plots.contour(
            x1, x2, (x, y) -> softmax(jem([x, y]))[target], 
            fill=true, alpha=0.5, title="Target: $target", cbar=true,
            xlims=xlims,
            ylims=ylims,
        )
        Plots.scatter!(X[1,:], X[2,:], color=Int.(labels.refs), group=Int.(labels.refs), alpha=0.5)
        Plots.scatter!(
            X̂[1,:], X̂[2,:], 
            color=repeat([target], size(X̂,2)), 
            group=repeat([target], size(X̂,2)), 
            shape=:star5, ms=10
        )
        push!(plts, plt)
    end
    plt = Plots.plot(plts..., layout=(1, 2), size=(2*500, 400))
    display(plt)
end
```

### 2023-12-13 paper/camera-ready/altmeyer.tex
acknowledgements

>>>>

\section*{Acknowledgements}

Some of the members of TU Delft were partially funded by ICAI AI for Fintech Research, an ING---TU Delft
collaboration. Research reported in this work was partially or completely facilitated by computational resources and support of the DelftBlue~\citep{DHPC2022} and the Delft AI Cluster (DAIC: https://doc.daic.tudelft.nl/) at TU Delft. The authors would like to thank Azza Ahmed, in particular, for her tremendous help with running Julia jobs on the cluster. The work remains the sole responsibility of the authors.
